---
title: "NYPD Shooting Incident Data Project"
author: "W. Jencyowski"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(naniar)
library(ggsankey)
# here is a test change for the markdown file
```

## Data Import
```{r data import}
# NYPD Shooting Incident Data (Historic) from www.catalog.data.gov
nypd_url <- "https://data.cityofnewyork.us/api/v3/views/833y-fsy8/export.csv?accessType=DOWNLOAD"
nypd_raw <- read_csv(nypd_url)
```

## Renaming
### Renaming each column for clarity and ease of reading
```{r renaming}
nypd <- nypd_raw |> 
    rename(
        Key = INCIDENT_KEY,
        Date = OCCUR_DATE,
        Time = OCCUR_TIME,
        Boro = BORO,
        Precinct = PRECINCT,
        Location = LOC_CLASSFCTN_DESC,
        Location_description = LOCATION_DESC,
        Murder = STATISTICAL_MURDER_FLAG,
        Perp_age = PERP_AGE_GROUP,
        Perp_sex = PERP_SEX,
        Perp_race = PERP_RACE,
        Vic_age = VIC_AGE_GROUP,
        Vic_race = VIC_RACE,
        Vic_sex = VIC_SEX,
        )
```

## Refactoring
### Refactoring column type based off the dataset footnotes 
```{r refactoring}
nypd$Key <- as.integer(nypd$Key)
nypd$Date <- as.Date(nypd$Date, "%m/%d/%Y")
# OCCUR_TIME is already is already hr/min/sec
nypd$Boro <- as.factor(nypd$Boro)
nypd$Precinct <- as.integer(nypd$Precinct)
nypd$Location <- as.factor(nypd$Location)
nypd$Location_description <- as.factor(nypd$Location_description)
```

## Cleaning perp and victim demographics
### This next section cleans the age, sex, and race columns using the naniar library, specifically the `replace_with_na` function.

#### Finding perp demo outliers
```{r perp demo outliers}
# Using the `unique()` command  to find each demegraphic that doesn't fit into the established convention
unique(nypd$Perp_age)
unique(nypd$Perp_sex)
unique(nypd$Perp_race)
```

#### Replacing perp demo outlier values with NA
```{r clean perp demo}
# entries in each column that don't match convention. Unknowns are changed to NA for uniformity across the data set
perp_age_filter <- c("(null)", "2021", "1028", "UNKNOWN", "1020", "940", "224")
perp_sex_filter <- c("(null)", "U")
perp_race_filter <- c("(null)", "UNKNOWN")

# replacing those values with NA
nypd <- nypd |> 
    replace_with_na(replace = list(Perp_age = perp_age_filter,
                                   Perp_sex = perp_sex_filter,
                                   Perp_race = perp_race_filter)) 
```

#### Finding vic demo outliers
```{r vic demo outliers}
# the same process as above is done for the victims
unique(nypd$Vic_age)
unique(nypd$Vic_sex)
unique(nypd$Vic_race)
```

#### Replacing vic demo outlier values with NA
```{r clean vic demo}
# creating a list of strings from each column that doesn't match convention
# unknowns are changed to NA for uniformity across the data set
vic_age_filter <- c("UNKNOWN", "1022")
vic_sex_filter <- c("U")
vic_race_filter <- c("UNKNOWN")

# replacing those values with NA
nypd <- nypd |> 
    replace_with_na(replace = list(Vic_age = vic_age_filter,
                                   Vic_sex = vic_sex_filter,
                                   Vic_race = vic_race_filter)) 
```

#### Mutating age demos to ordered factors
```{r}
nypd <- nypd |> 
    mutate(Perp_age = factor(Perp_age, levels = c("<18", "18-24", "25-44", "45-64", "65+")),
          Vic_age = factor(Vic_age, levels = c("<18", "18-24", "25-44", "45-64", "65+"))
    )
```


```{r}
race <-  nypd |>
    select(Key, Date, Boro, Perp_race, Vic_race) |> 
    mutate(Vic_race_ord = fct_infreq(Vic_race))
```

## Vic Race stacked bar plot
```{r stacked bar plot}
ggplot(race, aes(x = Boro, fill = Vic_race_ord)) +
    geom_bar(position = "dodge")
```


```{r basic line plot}
ggplot(race, aes(x = Date, y = Total)) +
    geom_line()
```