---
title: "NYPD Shooting Incident Data Project"
author: "W. Jencyowski"
output:
    html_document:
        toc: true
        toc_float: true
        theme: flatly
        highlight: tango
---

```{r setup, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(naniar)
```

# Data Import
```{r data import, cache = TRUE}
# NYPD Shooting Incident Data (Historic) from www.catalog.data.gov
nypd_url <- "https://data.cityofnewyork.us/api/v3/views/833y-fsy8/export.csv?accessType=DOWNLOAD"
nypd_raw <- read_csv(nypd_url)
```

# Tidying
### Renaming columns for clarity
```{r renaming}
nypd <- nypd_raw |> 
    rename(
        Key = INCIDENT_KEY,
        Date = OCCUR_DATE,
        Time = OCCUR_TIME,
        Boro = BORO,
        Vic_race = VIC_RACE,
        )
```

### Refactoring using the dataset footnotes 
```{r refactoring}
nypd$Key <- as.integer(nypd$Key)
nypd$Date <- as.Date(nypd$Date, "%m/%d/%Y")
# TIME is already factored as hr/min/sec
nypd$Boro <- as.factor(nypd$Boro)
```

### Cleaning victim demographics
##### This next section cleans the race columns using the naniar library, specifically the `replace_with_na` function.

#### Finding Victim race outliers
```{r vic demo outliers}
unique(nypd$Vic_race)
```

#### Replacing Victim race outlier values with NA
```{r clean vic demo}
# creating a string for races that don't match convention
# unknowns are changed to NA for uniformity across the data set
vic_race_filter <- c("UNKNOWN")

# replacing those "UNKNOWN"s with NA
nypd <- nypd |> 
    replace_with_na(replace = list(Vic_race = vic_race_filter)) 
```

# Visualizations

## Victim Race Bar Plots
#### The first set of bar plots show the count of all victims, separated by race, for all Boros. The Zoomed plot removes the three most frequent races in order to better observe the less represented races.
#### The second plot shows the proportion of victim race within each Boro.

### Victim frequency by Boro and race

```{r}
race <-  nypd |>
    select(Key, Date, Boro, Vic_race) |> 
    mutate(Vic_race_ord = fct_infreq(Vic_race))
```

```{r dodge bar, out.width = "100%"}
ggplot(race, aes(x = Boro, fill = Vic_race_ord)) +
    geom_bar(position = "dodge") +
    scale_fill_discrete(labels = function(x) stringr::str_wrap(x, width = 12)) +
    theme(legend.position = "right",
          legend.key.spacing.y = unit(0.5, "cm"),
          plot.title = element_text(hjust = 0.5)) +
    labs(y = "Incidents", 
         fill = "Victim Race")
```

### Victim frequency by Boro and race (Zoomed)

```{r dodge bar zoomed, out.width = "100%"}
race |> 
    filter(!Vic_race_ord %in% c("BLACK", "WHITE HISPANIC", "BLACK HISPANIC")) |> 
    ggplot(aes(x = Boro, fill = Vic_race_ord)) +
        geom_bar(position = "dodge") +
        scale_fill_discrete(drop = FALSE, labels = function(x) stringr::str_wrap(x,
            width = 12)) +
        theme(legend.position = "right",
            legend.key.spacing.y = unit(0.5, "cm"),
            plot.title = element_text(hjust = 0.5)) +
        labs(y = "Incidents", 
            fill = "Victim Race")
```


### Proportion of victim race within each Boro
```{r victim proportion, out.width = "100%"}
ggplot(race, aes(x = Boro, fill = Vic_race_ord)) +
    geom_bar(position = "fill") +
    scale_fill_discrete(labels = function(x) stringr::str_wrap(x, width = 12)) +
    theme(legend.position = "right",
          legend.key.spacing.y = unit(0.5, "cm"),
          plot.title = element_text(hjust = 0.5)) +
    labs(title = "Victim race proportion within each Boro",
         y = "Proportion", 
         fill = "Victim Race")
    
```

### Victim Race Analysis
#### This data shows that across each borrow, the proportion of victim race is roughly the same across the boros, however, Brooklyn is by and far the boro with the most shooting incident. 

## Day of Week vs. Time Incident Frequency Heat Map
#### The heat map will show the frquency of the incidents based off the time it occured and the day of the week. The 'Key' variable is needed to ensure each incident only gets counted once, since some Keys are listed more than once if there were multiple victims
```{r time v day of week}
# creating a new df 'dow_heat' with the Key, Date, and Time.
# added a new column that contains the day of the week
dow_heat <- nypd |> 
    select(Key, Date, Time) |> 
    mutate(
        # using the weekdays function to find the day of the week
        Weekday = weekdays(Date),
        # days of the week are ordered as such for visual clarity in the heat map
        Weekday = factor(Weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                                             "Friday", "Saturday", "Sunday")),
        # extracting just the hour of each incident
        Hour = hour(Time),
    ) |>
    # counting each incident as once occurence instead of how many victims there were
    distinct(Key, Weekday, Hour, .keep_all = TRUE) |>
    count(Weekday, Hour, name = "Incidents")
```

```{r heat map, out.width = "100%"}
# using geom_tile to create heat map of day of week vs time of shooting
ggplot(dow_heat, aes(x = Hour, y = fct_rev(Weekday), fill = Incidents)) +
    geom_tile(color = "white", lwd = 0.1, linetype = 1) +
    geom_text(aes(label = Incidents), color = "white", size = 3) +            
    scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20),
                     expand = c(0, 0)) +
    scale_y_discrete(expand = c(0,0)) +
    coord_fixed(ratio = 1) +
    labs(y = NULL, title = "Day of Week vs. Time Incident Frequency Heat Map") +
    theme(axis.text.y = element_text(vjust = 0.5),
          #axis.text.x = element_text(size = 12),
          panel.background = element_blank(),
          plot.background = element_blank(),
          plot.title = element_text(size = 18, hjust = 0.5),
          legend.position = "none")
```

### Heat Map Analysis
####  The data shows incidents most frequently occur during Friday and Saturday nights and into the morning, steeply falling off around 5 a.m. The incidents have a sharp decline in frquency during the weekday morning commute hours, Monday-Friday between 6 and 9 a.m.


# Modelled Data

### Incident Interval Calculation
```{r Incident Intervals}
intervals <- nypd |> 
    select(Key, Date, Time) |> 
    mutate(date_time = ymd_hms(paste(Date, Time))) |>  
    distinct(Key, .keep_all = TRUE) |> 
    arrange(date_time) |>
    mutate(interval_hours = as.numeric(difftime(date_time, lag(date_time), units = "hours"))) |> 
    filter(!is.na(interval_hours)) |> 
    ungroup()

# calculating the mean and lambda interval
mean_int <- mean(intervals$interval_hours)
lambda_int <- 1 / mean_int
```

### Incident Interval Model
#### The histogram is plotted using density rather than raw count to ensure the y-axis scale will be the same between the histogram and the density curve.
```{r histogram plot, out.width = "100%"}
ggplot(intervals, aes(x = interval_hours)) +
    # using the proportion density rather than actual count to overlay exponential distribution curve
geom_histogram(aes(y = after_stat(density)),
                       binwidth = 1,
                       fill = "#F8766D",
                       color = "white") +
                           # exponential curve
                        stat_function(fun = dexp,
                  args = list(rate = lambda_int),
                  color = "#00BFC4",
                  linewidth = 1) +
                      # vertical line showing the calculate mean
                    geom_vline(xintercept = mean_int,
               linetype = "dashed",
               color = "#00BA38",
               linewidth = 1.1) +
                   labs(title = "Shooting Interval Density",
                subtitle = "Plot truncated to 48 hours for visual clarity",
         x = "Incident Interval in Hours",
         y = "Density") +
             theme(plot.title = element_text(size = 18, hjust = 0.5),
            plot.subtitle = element_text(hjust = 0.5),
          panel.background = element_blank(),
          plot.background = element_blank()) +
              # adjusting the shown x-axis for clarity
            coord_cartesian(xlim = c(0, 48))
```

## Incident Interval Distribution Analysis
#### The intervals between incidents follows a Poisson/Exponential distribution. The vertical line represents the calculated interval mean, approximately 7.1 hours between each shooting incident

# Potential Bias in Data and Analysis
### Reporting
#### This dataset only records incidents where someone was actually wounded or killed. It does not include incidents where people were shot at and not wounded. It may seem obvious when said aloud, but this data doesn't include incidents that weren't reported. There may have been incidents where someone was wounded but the police weren't called.

### Racial
#### New York is one of the most diverse cities in the world. From a data standpoint, it may make sense to group a victim's race into a nice and neat category, but there may be victims who would potentially not identify with any of the given demographics, but were lumped into the closest fitting choice.

# Session Info
```{r session info}
sessionInfo()
```